\documentclass[letterpaper]{article}
\input{../../../preamble.tex}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}

\pagestyle{fancy}
\fancyhf{}
\rhead{CSE 131}
\chead{Friday, May 19, 2023}
\lhead{Lecture 21}
\rfoot{\thepage}

\setlength{\parindent}{0pt}

\begin{document}

\section{Optimization (Continued)}
This section continues the previous section.

\subsection{Optimization: Register Allocation}
Let's consider the following code: 
\begin{verbatim}
    (let (n (+ 5 9))
        (let (m (+ 2 3))
            (let (x (+ n 1))
                (let (y (+ m 2))
                    (+ x y)))))\end{verbatim}

The corresponding assembly\footnote{With tag checks removed to make the assembly more concise.}, along with the corresponding code from the above, is shown below.
\begin{verbatim}
    sub rsp, 40
      mov rax, 10
      mov [rsp + 0], rax    ; LHS of (+ 5 9)
      mov rax, 18
      add rax, [rsp + 0]

      mov [rsp + 0], rax    ; Variable n in (let (n ...))

      mov rax, 4
      mov [rsp + 8], rax    ; LHS of (+ 2 3)
      mov rax, 6
      add rax, [rsp + 8]

      mov [rsp + 8], rax    ; Variable m

      mov rax, [rsp + 0]    ; Variable n lookup 
      mov [rsp + 16], rax   ; LHS of (+ n 1)
      mov rax, 2
      add rax, [rsp + 16]

      mov [rsp + 16], rax   ; Variable x

      mov rax, [rsp + 8]    ; Variable m lookup 
      mov [rsp + 24], rax   ; LHS of (+ m 2)
      mov rax, 4
      add rax, [rsp + 24]

      mov [rsp + 24], rax   ; Variable y 

      mov rax, [rsp + 16]   ; Variable x lookup 
      mov [rsp + 32], rax
      mov rax, [rsp + 24]   ; Variable y lookup 
      add rax, [rsp + 32]
    add rsp, 40\end{verbatim}
One thing to notice immediately is that we reused some memory locations. One example is \code{[rsp + 8]}, which is where we stored both a temporary for addition and a value associated with a variable. We can generalize how many memory locations we ultimately \emph{will} use by using the \code{depth} function. In particular, if $\code{depth(expr)} \leq \text{Available Registers}$, then we can avoid memory entirely.

\bigskip 

There are two questions we should now consider.
\begin{enumerate}
    \item (\code{x86\_64}.) What registers should we use?
    \begin{mdframed}
        We can use the registers \code{rbx}, \code{r12}, \code{r13}, \code{r14}, which are callee-saved registers. Note that we aren't using \code{r15} because this register is specifically the heap pointer.
    \end{mdframed}
    \item (Design.) How should we implement this?
    \begin{mdframed}
        We can create a \code{Loc} \emph{enum} that holds either a register or a stack location (offset). Then, our environment can be represented by \code{HashMap<String, Loc>}.

        \bigskip 

        Suppose we have a list of registers that we can use. We can create a \code{get\_loc} function which takes a stack index and returns the new location to be used; this might look something like 
        \begin{verbatim}
    let regs = [...];
    get_loc(si):
        if si < regs.size():
            return regs[si];
        else:
            return Stack(si - regs.len());\end{verbatim}
        Then, we can use this location to update the environment, like 
        \begin{verbatim}
    ... 
    | ELet(x, val, body) => {
        env.update(x, get_loc(si));
    }\end{verbatim}
        Note that, while this is an \emph{improvement} to how our program is compiled, this can still be made a \emph{lot better}. Some other implementation notes to consider include: 
        \begin{itemize}
            \item We need to add code to save and restore registers in function definitions.
            \item We need to compute stack size based on \code{depth - available registers}.
        \end{itemize}
        Some improvements we could make to what we have so far include 
        \begin{itemize}
            \item Registers for outer bindings and stack for inner bindings. 
            \item Frequency matters. 
            \item Precompute registers and locations for all variables and temporaries across functions. 
            \item Are we using the minimal number of locations? (e.g., is the depth minimal?)
        \end{itemize}
    \end{mdframed} 
\end{enumerate}
\textbf{Remark:} The register allocation algorithm we're talking about, which uses an idea similar to \code{depth}, is similar to the \emph{Sethi-Ullman algorithm}.

\subsubsection{High Level Steps}
At a high level, we aim to answer the following questions:
\begin{itemize}
    \item The first step is to find the minimal number of locations needed to store all the working variables in an expression.
    \item What pairs of variables must be stored (or must be ``live'') at the same time? 
\end{itemize}

\end{document}