\documentclass[letterpaper]{article}
\input{../../../preamble.tex}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}

\pagestyle{fancy}
\fancyhf{}
\rhead{CSE 130}
\chead{Monday, May 16, 2022}
\lhead{Lecture 21}
\rfoot{\thepage}

\setlength{\parindent}{0pt}

\begin{document}
\section{Lexing and Parsing}
\subsection{Running the Parser}

Recall that there were two issues. 
\begin{itemize}
    \item Wrong Precedence
    \begin{verbatim}
    $ evalString [] "2 * 5 + 5"
    20\end{verbatim}
    This is invalid, and is due to the fact that our grammar is \textbf{ambiguous} -- there are multiple ways to parse the string \code{2 * 5 + 5}; one way is correct (\code{(2 * 5) + 5}) and one way is incorrect (\code{2 * (5 + 5)}). Thus, we want to tell \code{happy} that \code{*} has higher precedence than \code{+}.

    \item Wrong Associativity
    \begin{verbatim}
    $ evalString [] "2 - 1 - 1"
    2\end{verbatim}
    So, we also need to tell \code{happy} that \code{-} is \textbf{left-associative}.     
\end{itemize}
Therefore, we need to tell \code{happy} about precedence and associativity. 

\subsubsection{Solution 1: Grammar Factoring}
We can split the \code{AExpr} non-terminal into multiple ``levels.''
\begin{verbatim}
    Aexpr : Aexpr '+' Aexpr
          | Aexpr '-' Aexpr
          | Aexpr2

    Aexpr2 : Aexpr2 '*' Aexpr2
          | Aexpr2 '/' Aexpr2
          | Aexpr3

    Aexpr3 : TNUM
          | ID
          | '(' Aexpr ')'\end{verbatim}
Note that \code{AExpr} is the most general term. Then, \code{AExpr2} is slightly more specific; it only has expressions relating to multiplication and division, and numbers/identifiers. \code{AExpr3} is the most strict, with only numbers/identifiers. Intuitively, \code{AExpr2} will ``bind tighter'' than \code{AExpr}. We also have \code{AExpr3}, which is the ``tightest.''

\bigskip 

This fixes the issue with \code{2 * 5 + 5}, but fails to parse \code{5 + 5}.

\begin{mdframed}[]
    (Quiz.) With this new grammar, can we parse \code{2 - 1 - 1} the wrong way?

    \begin{enumerate}[(a)]
        \item Yes.
        \item No.
    \end{enumerate}

    \begin{mdframed}[]
        The answer is \textbf{B}. There are still multiple ways to parse \code{2 - 1 - 1}.
    \end{mdframed}
\end{mdframed}
How do we fix this? One way to do so is to disallow the right-hand side of a minus to be a minus. 

\begin{verbatim}
    Aexpr : Aexpr '+' Aexpr2
          | Aexpr '-' Aexpr2
          | Aexpr2

    Aexpr2 : Aexpr2 '*' Aexpr3
          | Aexpr2 '/' Aexpr3
          | Aexpr3

    Aexpr3 : TNUM
          | ID
          | '(' Aexpr ')'\end{verbatim}

\subsubsection{Solution 2: Parser Directives}
We can just use a special syntax for the parser generator; in \code{happy}, this is the syntax\footnote{Other parser generators may use other syntax.}: 
\begin{verbatim}
    %left '+' '-'
    %left '*' '/'\end{verbatim}
This means that 
\begin{itemize}
    \item All our operations are left-associative. 
    \item Operators on the lower line have higher precedence.
\end{itemize}
Note that operations like \emph{applications}, which is left-associative, do not have an operator; thus, you still need to worry about things like these. 




\end{document}