\documentclass[letterpaper]{article}
\input{../../../preamble.tex}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}

\pagestyle{fancy}
\fancyhf{}
\rhead{CSE 130}
\chead{Wednesday, March 30, 2022}
\lhead{Lecture 2}
\rfoot{\thepage}

\setlength{\parindent}{0pt}

\begin{document}

\section{The Lambda Calculus}
Most programming languages have modern features like 
\begin{itemize}
    \item Assignments
    \item Types
    \item Conditions
    \item Loops
    \item Classes
    \item And so on\dots
\end{itemize}

However the smallest universal language doesn't need \emph{any of these} at all.

\subsection{The Smallest Universal Language}
What is the smallest universal language?

\subsubsection{What is Computable?}
Before the 1930s, there was the informal notion of an \emph{effectively calculable} function. That is, it can be computed by a human with pen and paper, followed by an algorithm. 

\subsubsection{Formalization of a Language}
\textbf{Alan Turing} introduced the \emph{Turing Machine}, which is an infinite tape with some symbols, a head that can read from and write to the tape. To actually interact with a Turing machine, it has a state machine which has a bunch of states, along with a transition function, which tells it how to move around and what to do. The programming language is essentially the transition function of the Turing machine. 

\bigskip 

\textbf{Alonzo Church} came up with the \emph{Lambda Calculus}, which is (in some sense) simpler than a Turing machine. 

\bigskip 

\textbf{Peter Landin} used the Lambda Calculus to formalize the notion of a programming language. Lambda Calculus was influential in the creation of many modern programming languages, especially functional programming languages like Haskell. 

\subsubsection{The Lambda Calculus}
It has one feature: \emph{functions}. It does not have assignments, primitive types, control flow, recursion, etc. It literally only has \emph{functions}. Specifically, you can
\begin{itemize}
    \item define a function. 
    \item call a function. 
\end{itemize}

\subsubsection{Describing a Programming Language}
We're interested in two things: 
\begin{itemize}
    \item Syntax: what do programs look like? We use formal grammars (context-free grammars) to explain the syntax of a programming language.
    \item Semantics: what do programs mean? Specifically, \emph{operational semantics}, or the idea of how programs execute step-by-step. 
\end{itemize}

\subsection{Syntax}
We have one syntactical category: expressions (also called $\lambda$-terms).
\[\underbrace{\code{E}}_{\text{Expression}}\]
What can this expression expand to? 
\[\code{E ::= } \underbrace{\code{x}}_{\text{Variable}} \code{ | } \overbrace{\code{\bsttt{x} -> E}}^{\text{Abstraction}} \code{ | } \underbrace{\code{E1 E2}}_{\text{Application}}\]

The expression $E$ can be one of three kinds: 
\begin{itemize}
    \item \underline{Variables:} \code{x} can be any variable, e.g. \code{y} or \code{z} or even something like \code{apple}. We should think of variables as mathematical variables (immutable). It does not change its value over time; it's like a variable in math where all it does is holds its value. 
    
    \item \underline{Abstraction:} We can think of \code{\bsttt{x} -> E} in the following mathematical way: 
    \[f(x) = E\]
    This is, specifically, a \textbf{nameless function} that takes in input $x$ (i.e. formal parameter) and returns an expression $E$ (i.e. the body).
    
    \item \underline{Application:} Here, \code{E1} is the function and \code{E2} is the argument. We can think of this as \code{E1(E2)} in any programming language. In mathematics, we might write $f(5)$. In Lambda Calculus, this would be \code{f 5}. 
\end{itemize}
Note that \code{E}, \code{E1}, and \code{E2} can itself be a variable, abstraction, or application. Additionally, note that application binds stronger than abstractions\footnote{This will be important later on when we have syntactical sugar.}. 

\bigskip 

Effectively, in Lambda Calculus, functions are \emph{first-class}: we can apply functions to other functions, and a function can return another function. 

\subsection{Examples}
We will now go over some examples. 
\begin{itemize}
    \item Consider the following program. 
    \begin{verbatim}
        apple
    \end{verbatim}
    This does nothing, but is a syntactically value program in Lambda Calculus.

    \item Consider the following program. 
    \begin{verbatim}
        apple banana
    \end{verbatim}
    This is an \emph{application} of the variable \code{apple} to the variable \code{banana}.

    \item Consider the following program. 
    \begin{verbatim}
        \x -> x
    \end{verbatim}
    The identity function, which says that \emph{for any \code{x}, compute \code{x}}. This is the first program which is meaningful for us; it is a \emph{very} important function. 

    \item Consider the following program.
    \begin{verbatim}
        (\x -> x) apple
    \end{verbatim}
    This (whole program) is an application of the identity function to the variable \code{apple}. The result of this program will be \code{apple}. Note that \code{()} is not in the grammar that we described above; the grammar is known as an \emph{abstract syntax}, which simply describes what the programming language should have. In other words, the parentheses are ignored. Note that if we have 
    \begin{verbatim}
        \x -> x apple 
    \end{verbatim}
    which is a different program. Here, \code{x apple} is the body. 

    \item Consider the following program.
    \begin{verbatim}
        \x -> (\y -> y)
    \end{verbatim}
    Here, we introduce another variable \code{y}. This takes one argument \code{x}, completely ignores \code{x}, and returns the identity function. Comparing this to the previous example, all we're doing is changing the name of the formal parameter. 
    
    \item Consider the following program.
    \begin{verbatim}
        \f -> f (\x -> x)
    \end{verbatim}
    All this does is takes a argument \code{f}, and applies that argument to the identity function.
\end{itemize}


\subsection{Two Input Arguments}
Suppose you wanted a function that takes arguments \code{x} and \code{y} and returns \code{y}. How would you implement this? 

\bigskip 

Consider the following program.
\begin{verbatim}
    \x -> (\y -> y)
\end{verbatim}
Here, this function returns the identity function. This is the same thing as a function that takes two arguments and returns the second one. 

\subsubsection{Applying Function to Two Arguments}
For example, how do we apply \code{\bsttt{x} -> (\bsttt{y} -> y)} to \code{apple} and \code{banana}?

\bigskip 

We can do something like 
\begin{verbatim}
    ((\x -> (\y -> y)) apple) banana 
\end{verbatim}
This first applies \code{apple} and then applies to \code{banana}. 


\subsection{Syntactical Sugar}
The following are syntactical sugar for common operations. 

\begin{center}
    \begin{tabular}{p{3in} | p{3in}}
        \textbf{Instead of} & \textbf{We can write} \\ 
        \hline 
        \code{\bsttt{x} -> (\bsttt{y} -> (\bsttt{z} -> E))} & \code{\bsttt{x} -> \bsttt{y} -> \bsttt{z} -> E} \\ 
        \code{\bsttt{x} -> \bsttt{y} -> \bsttt{z} -> E} & \code{\bsttt{x} y z -> E} \\ 
        \code{(((E1 E2) E3) E4)} & \code{E1 E2 E3 E4}
    \end{tabular}
\end{center}
Additionally, the following are equivalent: 
\begin{verbatim}
    \a -> (\b -> b)
    \a -> \b -> b
    \a b -> b
\end{verbatim}
The parentheses are applied starting from the right side.

\end{document}