\documentclass[letterpaper]{article}
\input{../../../preamble.tex}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}

\pagestyle{fancy}
\fancyhf{}
\rhead{Math 103B}
\chead{Wednesday, February 09, 2022}
\lhead{Lecture 16}
\rfoot{\thepage}

\setlength{\parindent}{0pt}

\begin{document}

\section{More on Irreducible Polynomials}
We will continue our discussion on irreducible polynomials.

\subsection{Eisenstein's Criterion}
\begin{theorem}{}{}
    Let 
    \[f(x) = a_n x^n + a_{n - 1} x^{n - 1} + \dots + a_0 \in \Z[x]\]
    If there is a prime $p$ such that $p \nmid a_n$, $p \mid a_{n - 1}, \ldots, p \mid a_0$ and $p^2 \nmid a_0$, then $f(x)$ is irreducible over $\Q$.  
\end{theorem}

\begin{mdframed}[]
    Suppose, towards a contradiction, that $f(x)$ is reducible over $\Q$. This implies that $f(x)$ is reducible over $\Z$. We write 
    \[f(x) = g(x) h(x)\]
    with $0 < \deg g(x), \deg h(x) < \deg f(x)$. We then write 
    \[g(x) = b_r x^r + \ldots + b_1 x + b_0\]
    \[h(x) = c_s x^s + \ldots + c_1 x + c_0\]
    It follows that 
    \[f(x) = g(x) h(x) = b_0 c_0 + (b_1 c_0 + b_0 c_1) x + \ldots\]
    If $p | a = b_0 c_0$, then it follows that $p | b_0$ or $p | c_0$. We also know that $p \nmid a$, implying that $p^2 \nmid b_0$ \emph{and} $p^2 \nmid c_0$. Suppose WLOG that $p | b_0$ and $p \nmid c_0$. Suppose this holds for the leading term, $p \nmid a_n = b_r c_s$; thus, $p \nmid b_r$ and $p \nmid c_s$. Let $t$ be the minimal index such that $p \nmid b_t$ ($t < r < n = \deg f(x)$). Then, $p | a_t = b_t c_0 + b_{t - 1}c_1 + \ldots b_0 c_t$. Because $t$ is minimal, it follows that $p | b_{t - 1} \ldots p | b_0$ all holds. This implies that 
    \[0 \equiv b_t c_0 + 0 \Mod{p}\]
    but this is a contradiction since $b_0 \not\equiv 0 \Mod{p}$ and $c_0 \not\equiv 0 \Mod{p}$, a contradiction as $\Z / p\Z$ is an integral domain. 
\end{mdframed}

\subsubsection{Cyclotomic Polynomial}
Consider the $p$th Cyclotomic polynomial. That is, for any prime $p$, the $p$th Cyclotomic polynomial
\[\Phi_{p}(x) = \frac{x^p - 1}{x - 1} = x^{p - 1} + x^{p - 2} + \ldots + x + 1\]
is irreducible over $\Q$. 

\subsubsection{Example: Polynomial}
Is the polynomial
\[x^5 + 2x^2 + 2\]
irreducible?

\begin{mdframed}[]
    Consider the prime $p = 2$. It's obvious that, for each coefficient, $2 \nmid 1$ is true, $2 | 2$ is true, and $2 | 2$ is true. Additionally, $4 \nmid 2$, so by Eisenstein's Criterion, it follows that this polynomial is irreducible.
\end{mdframed}

\subsection{Maximal Ideals}
\begin{theorem}{}{}
    Let $\F$ be a field and $p(x) \in F[x]$. Then, $\cyclic{p(x)}$ is maximal in $\F[x]$ if and only if $p(x)$ is irreducible over $F$. 
\end{theorem}

\begin{theorem}{}{}
    Let $\F$ be a field and $p(x)$ be an irreducible polynomial over $\F$. Then, $\F[x] / \cyclic{p(x)}$ is a field. 
\end{theorem}

\begin{corollary}{}{}
    Let $\F$ be a field and let $p(x), a(x), b(x) \in \F[x]$. If $p(x)$ is irreducible over $\F$ and $p(x) | a(x) b(x)$, then $p(x) | a(x)$ or $p(x) | b(x)$. 
\end{corollary}

\end{document}